---
title: "STATS 250 Lab 10"
subtitle: "Confidence Intervals and Hypothesis Tests for Proportions"
author: "Nick Seewald<br />nseewald&#64;umich.edu"
date: "Week of 11/2/2020"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
    css: ["xaringan-themer.css", "extras.css"]
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo = TRUE, include = TRUE, eval = TRUE, comment = NA, fig.height = 5, fig.width = 9)
source("plotNorm.R")
source("prop_test.R")
source("plot_ci.R")
par(mar = c(5, 5.5, 4, 2) + 0.1)
set.seed(1875)
```

```{r xaringan-themer, include = FALSE, warning = FALSE}
library(xaringanthemer)
# style_mono_light(base_color = "#00274c",
#                  header_font_google = google_font("DM Serif Display"),
#                  text_font_google = google_font("PT Sans", "400", "400i")
# )
style_duo(primary_color = "#00274c",
          secondary_color = "#ffcb05",
          header_font_google = google_font("DM Serif Display"),
          text_font_google = google_font("PT Sans", "400", "400i", "600"),
          code_font_google = google_font("Roboto Mono", "400"),
          text_font_size = "22pt"
)
```

```{r xaringanExtra, echo=FALSE}
xaringanExtra::use_webcam()
xaringanExtra::use_fit_screen()
xaringanExtra::use_editable()
xaringanExtra::use_panelset()
xaringanExtra::use_tachyons()
```

class: center middle

# Reminders `r emo::ji("bulb")`

Your tasks for the week running Friday 10/30 - Friday 11/6:

| Task | Due Date | Submission |
|:-----|:---------|:-----------|
| Vote (if eligible) | Tuesday 11/3 8:00PM ET | Your Election Precinct
| M-Write 2 Initial Submission | Thursday 11/5 4:59PM ET| Canvas
| Lab 10 | Friday 11/6 8:00AM ET | Canvas |
| Homework 7 | Friday 11/6 8:00AM ET | course.work |

---
# Lab Demo: ISRS Problem 3.9

.bg-washed-yellow.b--yellow.navy.ba.bw2.ph4[
.b[Life after college.] We’re interested in estimating the proportion of graduates at a mid-sized university who found a job within one year of completing their undergraduate degree. Suppose we conduct a survey and find out that 348 of the 400 randomly sampled graduates found jobs. The graduating class under consideration included over 4500 students.
]

---
# Part 1: What are we trying to find? What do we know?

.bg-washed-yellow.b--yellow.navy.ba.bw2.ph4[
.b[Life after college.] We’re interested in estimating the proportion of graduates at a mid-sized university who found a job within one year of completing their undergraduate degree. Suppose we conduct a survey and find out that 348 of the 400 randomly sampled graduates found jobs. The graduating class under consideration included over 4500 students.
]

--
### What is the population parameter of interest?

--

We want to find $p$, the proportion of **all** graduates at a mid-sized university who found a job within one year of completing their undergraduate degree.

---
# Lab Demo: ISRS Problem 3.9

.bg-washed-yellow.b--yellow.navy.ba.bw2.ph4[
.b[Life after college.] We’re interested in estimating the proportion of graduates at a mid-sized university who found a job within one year of completing their undergraduate degree. Suppose we conduct a survey and find out that 348 of the 400 randomly sampled graduates found jobs. The graduating class under consideration included over 4500 students.
]

--
### What is our *point estimate* of $p$?

--
$$\hat{p} = \frac{348}{400} = 0.87$$

---
# Part 2: Check Conditions

Before we can make a confidence interval using the normal distribution, we want to make sure that our data meet certain conditions.

### What conditions do we need to check?
--

1. **Independent observations:** graduates in the sample can't be related to each other
2. **Large enough sample:** $np\geq 10$ and $n(1-p) \geq 10$ (at least 10 "successes" and 10 "failures")

---
# Part 2: Check Conditions

### Check Independence
--

Our sample size of 400 is less than 10% of the population size of 4500. `r emo::ji("check")`

--
### Check sample size
--

We don't know $p$, so we'll check this condition with $\hat{p}$, our best guess of $p$:
$$n\hat{p} = 400\times 0.87 = \mathbf{348} \geq 10$$
$$n(1-\hat{p}) = 400 \times 0.13 = \mathbf{52} \geq 10$$
Both are at least 10 `r emo::ji("check")`

---
# Step 3: Compute a confidence interval

.bg-washed-yellow.b--yellow.navy.ba.bw2.ph4[
Calculate a 95% confidence interval for $p$, the proportion of graduates who found a job within one year of completing their undergraduate degree at this university, and interpret it in the context of the data.
]

Remember that a confidence interval generally looks like
$$\mathrm{estimate} \pm (\text{a few}) \times \mathrm{SE}_{\text{estimate}}$$

---
# Step 3: Compute a confidence interval

$$\mathrm{estimate} \pm (\text{a few}) \times \mathrm{SE}_{\text{estimate}}$$

Using a multiplier of 1.96 will give us a 95% confidence interval:

```{r echo = F, fig.align = "center"}
plotNorm(shadeValues = c(-1.96, 1.96), direction = "inside", cex = 2)
```

---
# Step 3: Compute a confidence interval

We know from section 3.1 that 
$$\mathrm{SE}_{\hat{p}} = \sqrt{\frac{p(1-p)}{n}}$$
but since we don't know $p$, we'll use $\hat{p}$. 

Use R as a calculator to compute $\mathrm{SE}_{\hat{p}}$, using $\hat{p} = 0.87$. 

--
```{r se-phat}
se <- sqrt(0.87 * (1 - 0.87) / 400)
se
```

---
# Step 3: Compute a confidence interval

Now let's compute the **margin of error**: the term that's added to and subtracted from the estimate to get the limits of the confidence interval.

$$\mathrm{estimate} \pm \underbrace{(\text{a few}) \times \mathrm{SE}_{\text{estimate}}}_{\text{margin of error}}$$
Remember that "a few" here means 1.96 (for a 95% confidence interval)

Use R as a calculator to compute the margin of error. 

--
```{r moe}
moe <- 1.96 * se
moe
```

---
# Step 3: Compute a Confidence Interval 

Our confidence interval, therefore, is 
$$0.87 \pm 0.033.$$
or
$$(0.837, 0.903)$$

### How do we interpret this confidence interval?
--

We are 95% confident that the population proportion of graduates at a mid-sized university who found a job within one year of completing their undergraduate degree is between .837 and .903. 

---
# Step 4: Interpreting a Confidence Level

.bg-washed-yellow.b--yellow.navy.ba.bw2.ph4[
What does "95% confidence" mean?
]

- **Imagine** that we know $p$ is 0.85.
- Take repeated samples from this population, and make a confidence interval using each sample
- We expect about 95% of the resulting confidence intervals to contain $p$ = 0.85

---
# Step 4: Interpreting a Confidence Level 

.pull-left[
```{r makeCIs, fig.show='hold'}
set.seed(5902)

# LINE ~120 OR SO
ci <- replicate(50, {
  s <- sample(0:1, size = 400,  #<<
              replace = TRUE, #<<
              prob = c(0.15, 0.85)) #<<
  
  pHat <- sum(s) / 400
  se <- sqrt(pHat * (1 - pHat) / 400)
  marginOfError <- 1.96 * se
  
  lowerLimit <- pHat - marginOfError
  upperLimit <- pHat + marginOfError
  
  c(lowerLimit, upperLimit) #<<
})

ci <- t(ci)
```
]

.pull-right[
```{r CIs}
head(ci)
```
]

---
# Step 4: Interpreting a Confidence Level

48/50 = 96% of the intervals contain $p=0.85$.

```{r plot_ci, fig.align = "center"}
plot_ci(lo = ci[, 1], hi = ci[, 2], m = 0.85)
```

---
# Step 4: Interpreting a Confidence Level

### How would you interpret the 95% confidence level?

--
If we repeated our sampling procedure many times, we would expect 95% of our resulting 95% confidence intervals to contain $p$, the true proportion of graduates who get a job within one year of finishing their undergraduate degrees.

---
# R can do this for us (line ~156)

We can have R make confidence intervals for us:

```{r prop_test1}
prop_test(x = 348, n = 400, conf.level = 0.95)
```

---
# Switch it up: 99% CI (line ~165)

Modify the code below to make a 99% confidence interval instead.

```{r prop_test1, eval = F}
```

--

```{r prop_test99}
prop_test(x = 348, n = 400, conf.level = 0.99)
```

How does the width of this interval compare to the 95% CI?

---
# Hypothesis Testing with `prop_test()`

`prop_test()` creates a confidence interval **and** performs a hypothesis test. Let's test the following hypotheses:

$$H_0: \ p = 0.5 \quad \text{vs.} \quad H_a: \ p < 0.5$$

```{r prop_testHT}
prop_test(x = 348, n = 400, 
          p = 0.5, alternative = "less")
```

---
# Hypothesis Testing with `prop_test()`

.pull-left[
```{r prop_testHT}
```
]
.pull-right[
```{r echo = F}
plotNorm(shadeValues = 14.8, xlim = c(-3, 15), direction = "less")
```
]

--

### Why is that p-value 1? 

--
We're testing to see if $p < 0.5$, but our data have $\hat{p} = 0.87$! Our data provide almost no evidence that $p < 0.5$, so we get a high p-value.

---
# Careful with `alternative`!

.pull-left[
```{r prop_test1}
```
]
.pull-right[
```{r prop_testHT}
```
]

If you want to make a confidence interval, you *must* do a two-sided test. Set `alternative = "two.sided"` or leave it blank.

---
# `prop_test()` for Two Proportions

Pass a **vector** of the numbers of successes `x` and a **vector** of sample sizes `n`.

.pull-left[
| | Successes | Failures | Total |
|-|-----------|----------|-------|
| Group 1 | 28 | 2 | 30 |
| Group 2 | 34 | 16 | 50 |
| Total | 62 | 18 | 80 |
]

.pull-right[
```{r propTest2Prop}
prop_test(x = c(28, 34),
          n = c(30, 50),
          conf.level = 0.9)
```
]

---
class: inverse
# Code Cheat Sheet `r emo::ji("computer")`
### `pnorm(q, mean = 0, sd = 1, lower.tail = TRUE)`
- **`q`** refers to the value you want to find the area above or below
  - `pnorm(q, 0, 1)` gives $P(Z < q)$ where $Z$ is $N(0,1)$
- **`mean`** refers to $\mu$, defaults to 0
- **`sd`** refers to $\sigma$, defaults to 1
- **`lower.tail`** controls which direction to "shade": `lower.tail = TRUE` goes less than `q`, `lower.tail = FALSE` goes greater than `q`; defaults to `TRUE`

---
class: inverse
# Code Cheat Sheet `r emo::ji("computer")`
### `qnorm(p, mean = 0, sd = 1, lower.tail = TRUE)`
- **`p`** refers to the area under the curve
  - `qnorm(p, 0, 1)` is the number such that the area to the left of it is `p`
- **`mean`** refers to $\mu$, defaults to 0
- **`sd`** refers to $\sigma$, defaults to 1
- **`lower.tail`** controls which direction to "shade": `lower.tail = TRUE` goes less than `q`, `lower.tail = FALSE` goes greater than `q`; defaults to `TRUE`

---
class: inverse
# Code Cheat Sheet `r emo::ji("computer")`
### `plotNorm(mean = 0, sd = 1, shadeValues, direction, col.shade, ...)`
- **`mean`** refers to $\mu$, defaults to 0
- **`sd`** refers to $\sigma$, defaults to 1
- **`shadeValues`** is a vector of up to 2 numbers that define the region you want to shade
- **`direction`** can be one of `less`, `greater`, `outside`, or `inside`, and controls the direction of shading between `shadeValues`. Must be `less` or `greater` if `shadeValues` has only one element; `outside` or `inside` if two
- **`col.shade`** controls the color of the shaded region, defaults to `"cornflowerblue"`
- **`...`** lets you specify other graphical parameters to control the appearance of the normal curve (e.g., `lwd`, `lty`, `col`, etc.)

---
class: inverse
# Code Cheat Sheet `r emo::ji("computer")`
### `prop_test(x, n, p = NULL, alternative = c("two.sided", "less", "greater"), conf.level = 0.95)`
- **`x`** is a vector of numbers of successes
- **`n`** is a vector of sample sizes
- **`p`** is is the null hypothesis value of $p$ or the hypothesized difference in proportions
- **`alternative`** can be one of `less`, `greater`, or `two.sided`, and controls the direction of the alternative hypothesis. Defaults to `two.sided`, which must be used to make a confidence interval
- **`conf.level`** controls the confidence level used to make the confidence interval, must be a single number between 0 and 1.
---
class: inverse
# Lab Project `r emo::ji("keyboard")`

.pull-left[
### Your tasks
- Complete the "Try It!" and "Dive Deeper" portions of the lab assignment by copy/pasting and modifying appropriate code from earlier in the document.
]

.pull-right[
### How to get help
- Use the "lab" tag on Piazza
- Email your lab instructor
]

---

class: center middle

# Reminders `r emo::ji("bulb")`

Your tasks for the week running Friday 10/30 - Friday 11/6:

| Task | Due Date | Submission |
|:-----|:---------|:-----------|
| M-Write 2 Initial Submission | Thursday 11/5 4:59PM ET| Canvas
| Lab 10 | Friday 11/6 8:00AM ET | Canvas |
| Homework 7 | Friday 11/6 8:00AM ET | course.work |

